Schema Evolution in Interactive Programming Systems

### 2024-11-06 sources/paper.tex
Cleanup, final tweaks, sources

\authorrunning{J. Edwards, T. Petricek, T. van der Storm, G. Litt}

@@

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Fig.~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referencable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a distinct type of entity that will be associated with the first one.

@@

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging a data edit
(Fig.~\ref{fig:conf-add}) with a schema and data edit (Fig.~\ref{fig:conf-tab}); the second involves
merging a schema and data edit (Fig.~\ref{fig:conf-tab}) with edit adding code (Fig.~\ref{fig:conf-budget}).
Formulas in the final version (Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
\label{fig:conf}
\end{figure}

@@

\begin{figure}[t]
\begin{adjustwidth}{-3em}{-3em}
\begin{minipage}{0.4\textwidth}
\includegraphics[width=\textwidth]{figures/metamodel.pdf}
\end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.4\textwidth}
  \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
  \end{minipage}
\hspace*{2pt}
\vline
\begin{minipage}{0.3\textwidth}
  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none]
machine Doors
var isClosed: bool = true

@@

\paragraph{Data Wrangling} The evolution of data formats is a concern in Data Science. Data~Diff~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation.
Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
\citet{petersohn20} present an algebra of operations on Python dataframes that include shape transformations.

>>>>

@@

\challenge{Extract Entity}
Acme needs to evolve the above schema and data into two tables (Figure~\ref{fig:db-customers}, \ref{fig:db-orders-link}),
one with orders that links to one with customers.  We refer to this as the \emph{Extract Entity}
operation, because it introduces a new \emph{entity}, a referencable holder of mutable attributes.
The operation is needed when we realize that some attributes of one type of entity should belong to
a distinct type of entity that will be associated with the~first~one.

@@

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging a data edit
(Figure~\ref{fig:conf-add}) with a schema and data edit (Figure~\ref{fig:conf-tab}); the second involves
merging a schema and data edit (Figure~\ref{fig:conf-tab}) with edit adding code (Figure~\ref{fig:conf-budget}).
Formulas in the final version (Figure~\ref{fig:conf-finfin}) are updated to match the new structure.}
\label{fig:conf}
\end{figure}

@@

\begin{figure}[t]
\begin{adjustwidth}{-2.5em}{-2.5em}
\begin{minipage}{0.4\textwidth}
\includegraphics[width=\textwidth]{figures/metamodel.pdf}
\end{minipage}
\vline
\begin{minipage}{0.4\textwidth}
  \includegraphics[width=\textwidth]{figures/doors-ast.pdf}
  \end{minipage}
\vline
\begin{minipage}{0.3\textwidth}
  \begin{lstlisting}[language=java,morekeywords={machine,on,state,var},basicstyle=\footnotesize,numbers=none,xleftmargin=1.5em]
machine Doors
var isClosed: bool = true

@@

\paragraph{Data Wrangling} The evolution of data formats is a concern in Data Science. The Data~Diff tool~\cite{sutton18} compares two versions of tabular data and finds a minimal matching transformation. AI~assistants~\cite{AIassistants} provide a mechanism for interactively correcting
transformations inferred by Data~Diff and similar automatic tools.
Wrangler~\cite{kandel11} observes interactive manipulation of sampled data to write data transformation programs.
\citet{petersohn20} present an algebra of operations on Python dataframes that include shape transformations.
